Unifying intra- and inter-specific variation in tropical tree mortality

### 2018-07-20 ms/manuscript.Rnw
More edits to improve readability

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of growth-related and growth-independent hazards on the overall mortality rates. We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and tolerance to shade outperform alternative formulations. Moreover, we show that while these three traits do affect both growth-dependent and growth-independent hazards, they account for only a small fraction of the total variability observed among species. Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests. It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

\dropcap{R}ates of plant mortality are known to vary widely among individuals within a species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}. This variation has considerable consequences for forest structure and dynamics. For example, the death of a single large tree can transfer up to 20000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz} and creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun. In models of forest dynamics, variation in tree mortality rates has been shown to have significantly larger impacts on forest structure relative to variation in absolute growth rates \cite{Johnson-2016}. Improving our understanding of the mortality process is therefore a priority for making accurate predictions about population, carbon and nutrient dynamics of forests; especially in an era of increased climate-driven forest die-off \cite{Anderegg:2012ep}.

@@

@@

Plants die via many causes, however, these broadly fall into two main categories \cite{Hawkes:2000ib,McDowell:2011dr}. The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair. The second are carbon-independent hazards, where plants die because of chance events that are unrelated to their growth rate, such as windfall or fire. An individual's total hazard at a given time point is therefore the sum of these two independent hazards. However, to date, statistical models of plant mortality have not attempted to simulataneously estimate these hazards from standard survival data, let alone examine how intra- and inter-specific factors affect them.

\begin{figure}[!hb]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
B) We consider three alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard; and 3) a function that combines both baseline and growth-dependent hazards. The parameters of the models are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote, or baseline hazard. Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g. windfall, fire) that kill a plant, irrespective of its growth rate. For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}).
D) The predictive skill of alternative models was evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing (1 fold; orange) and training (9 folds; green) data. Predictive capacity was assessed by averaging the log loss's obtained from the 10 test data predictions.}
\label{fig1}
\end{figure}

Here, we attempt to develop a empirical model capable of reconciling these conflicting growth-mortality relationships observed among individuals and between species while also explicitly estimating the independent contributions of carbon-dependent and independent hazards. We achieve this by evaluating the following, unified hazard function that incorporates individual-, species-, and temporal effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s[i]}}^{\text{Growth independent}} + \overbrace{\alpha_{s[i]} \, \textrm{e}^{-\beta_{s[i]} \, X_i(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for the independent effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$. Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$. Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), shade-intolerance ($\upsilon$) and maximum dbh ($\psi$). We also include species random effects that captures any remaining species-level differences that are not otherwise accounted for by species traits (for details see Methods).

To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard, and then compare their skill in predicting the observed mortality for \Sexpr{n_inds(BCI_training_full)} individual from \Sexpr{n_spp(BCI_training_full)} tree species at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken approximately at 5-year intervals over a 15 yr period (Fig. \ref{fig1}A). In total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models. We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across seven iterative stages of model development.

In stage 1, we ask whether mortality rates vary substantially among censuses, and thus, whether a census effect is required. We achieved this by fitting a constant that is invariant with respect to species, growth rate, or census period (i.e. the null model) and compared it to the same model but with a census random effect.
In stage 2, we assess whether a function incorporating both growth-dependent and growth-independent hazards outperform functions that only account for one or the other of these. Species effects were excluded, so we are simply asking which of three functional forms for $\lambda_i(t)$ (Fig. \ref{fig1}B) best predicts the survival data. The simplest form (Fig. \ref{fig1}B, left) assumes a constant growth-independent hazard rate. The second form (Fig. \ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth increases. The third form (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\textgreater 0$. Combined, these three models capture a variety of functional responses previously proposed (e.g. \cite{Kobe:1995tw, Wyckoff:2000un}), including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared. We also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
In stage 3, we examine whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was on growth-dependent, growth-independent, or both hazards. Finally, we fit a model that allowed parameters to vary by species-level variation that was otherwise not captured by wood density. This allowed us to ask what proportion of inter-specific hazard variability is explained by wood density. Using the final ``best'' model we also conducted post-hoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.

@@

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a small, but significant, increase in predictive skill (Fig. \ref{fig2}A). Individual census effects can be found in Table S1.

\begin{figure*}[!ht]
\includegraphics{figures/fig2.pdf}
\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% credible intervals) logarithmic loss, with lower values implying greater predictive skill.
A) Five sequential stages of model selection with each stage increasing in complexity. Null: constant ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate. This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
B) Predictive skill for wood density parameter combinations. Shading represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth. For both panels symbols represent functional form: null (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).}
\label{fig2}
\end{figure*}

\subsection*{Hazard functions}
Comparing the three hazard functions in Fig. \ref{fig1}B, we found that the third function --- with both growth-dependent and growth-independent terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the growth-independent only (i.e. null) or growth-dependent only functions (Fig. \ref{fig2}A). Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig. \ref{fig2}A). A summary of hyper parameter estimates can be found in Table S1.

\begin{figure*}[h!]
\includegraphics{figures/fig3.pdf}
\caption{Relationship between species growth-independent mortality rate and wood density. Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study. Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship.}
\label{fig3}
\end{figure*}

>>>>

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified a variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of growth-related and growth-independent hazards on the overall mortality rates. We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and tolerance to shade outperform alternative formulations. Moreover, we show that while these three traits do affect both growth-dependent and growth-independent hazards, they account for only a small fraction of the total variability observed among species. Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests. It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

\dropcap{R}ates of plant mortality are known to vary widely among individuals within a species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}. This variation has considerable consequences for forest structure and dynamics. For example, the death of a single large tree can transfer up to 20,000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz} and creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun. In models of forest dynamics, variation in tree mortality rates has been shown to have significantly larger impacts on forest structure relative to variation in absolute growth rates \cite{Johnson-2016}. Improving our understanding of the mortality process is therefore a priority for making accurate predictions about population, carbon and nutrient dynamics of forests; especially in an era of increased climate-driven forest die-off \cite{Anderegg:2012ep}.

@@

Plants die because of many causes, however, these broadly fall into two main categories \cite{Hawkes:2000ib,McDowell:2011dr}. The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair. The second are carbon-independent hazards, where plants die because of chance events that are unrelated to their carbon budget, such as windfall or fire. An individual's total hazard at a given time point is therefore the sum of these two independent hazards. However, to date, most statistical models of tree mortality do not explicitly estimate the independent effects of these two fundamanetal hazards faced by plants. Instead, plant mortality is typically modelled using: 1) logistic regression (i.e. with logit link), which does not estimate a rate, but rather the log odds of death at a given time point; or 2) via a proportional hazards approach such as a Cox proportional-hazard model or complementary-log-log link binomial regression, both of which assume hazards are multiplicative, and therefore, not independent of one another \cite{Kleinbaum:2012tn}.

There are several distinct advantages of modelling tree mortality as a function of two additive, and thus independent hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested.

@@

\begin{figure}[!hb]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
B) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (i.e. average hazard), 2) a growth-dependent hazard; and 3) a function that combines both growth dependent and independent hazards. The parameters of the models are biologically interpretable: $\alpha$ defines the the low growth rate effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote (Panel B; right), or baseline hazard (Panel B; left). In the most complex model (Panel B; right), $\alpha$ and $\beta$ capture growth-dependent hazards, while $\gamma$ captures growth-independent hazards (e.g. windfall, fire) that kill a plant, irrespective of its growth rate. For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}).
D) The predictive skill of alternative models was evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing (1 fold; orange) and training (9 folds; green) data. Predictive capacity was assessed by averaging the log loss's obtained from the 10 test data predictions.}
\label{fig1}
\end{figure}

Here, we develop a tree mortality model that not only reconciles the conflicting growth-mortality relationships observed among individuals and between species, but also explicitly accounts for carbon-dependent and carbon-independent hazards. We achieve this by evaluating the following unified hazard function that incorporates individual-, species-, and temporal effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s[i]}}^{\text{Carbon independent}} + \overbrace{\alpha_{s[i]} \, \textrm{e}^{-\beta_{s[i]} \, X_i(t)}}^{\text{Carbon dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for the additive, thus independent, effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$. Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$. Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), light requirement ($\upsilon$) and maximum dbh ($\psi$). We also include species random effects in order to capture species-level differences that are not otherwise accounted for by species traits (for details see Methods).

To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard (i.e. a null model) and 2) a growth-dependent only hazard. We then compare each functions ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken approximately at 5-year intervals over a 15 yr period (Fig. \ref{fig1}A). In total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models. We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across six iterative stages of model development.

In stage 1, we first ask whether average mortality rates vary substantially among censuses, and thus, whether a census effect is required. We achieved this by fitting a constant, $\gamma$, that is invariant with respect to species, growth rate, or census period (i.e. the null model) and compared it to a constant that is scaled by a census random effect, $\gamma\delta_{t}$. This stage revealed that there was significant variation in mortality rates observed among censuses, and as such a random effect for census was included as a scaling factor for all subsequent models.

In stage 2, we assessed the predictive performance of three alternative hazard functions (Fig. \ref{fig1}B). The simplest form (Fig. \ref{fig1}B, left) assumes a constant hazard. The second form (Fig. \ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth increases. The third form (Fig. \ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike the second functional form examined here, asymptotes at some baseline hazard $\textgreater 0$. Combined, these three models capture a variety of functional responses that are both used in empirical studies of plant mortality \cite{Kobe:1995tw,Wyckoff:2000un} and are represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, but which have not previously been systematically compared. In doing this, we also also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.

In the remaining stages, we then take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits, wood density, light demand and maximum dbh, affected model predictive performance. First, we individually fit an effect of each trait on all parameters (stage 3), we then examine 2-trait combinations on all parameters (stage 4) as well as the effect of three traits on all parameters (stage 5). Lastly, we fit a model that had included a species random effect on each parameter (stage 6). This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by these three traits. Because these models were computationally expensive to run (some requiring over 400 computing hours), we were unable to examine interactions among traits, or determine the optimal combination of traits with parameters.

@@

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates declining over time. The highest proportion of tree death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding the term $\delta$) we observed a small increase in predictive skill (Fig. \ref{fig2}A). Individual census effects can be found in Table S1.

\begin{figure*}[!ht]
\includegraphics{figures/fig2.pdf}
\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill.
A) Six sequential stages of model selection with each stage increasing in complexity. \textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate. This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters. Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).}
\label{fig2}
\end{figure*}

\subsection*{Hazard functions}
Comparing the three hazard functions summarised in Fig. \ref{fig1}B, we found that the third function --- the function containing both growth-dependent and growth-independent hazard terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline hazard or growth-dependent only functions (Fig. \ref{fig2}). Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig. \ref{fig2}A). A summary of hyper parameter estimates can be found in Table S1.

\begin{figure*}[h!]
\includegraphics{figures/fig3.pdf}
\caption{Estimated relationships between Relationship between species growth-independent mortality rate and wood density. Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study. Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship.}
\label{fig3}
\end{figure*}
